---
title: "Damage Caused by Severe Weather Events"
author: "Me"
date: "2025-08-11"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

The purpose of this analysis was to investigate severe weather events and determine which ones caused the most damage in terms of both human health and finances. First, I imported the raw data from the database website. The events in the database start in the year 1950 and end in November 2011. Then, I sorted the data into groups; first by total fatalities, then by total property damage. Finally, I graphed the two data groups into bar plots.

## Data Processing

First, I imported necessary libraries and downloaded the data. Then, I read in the data using the readr package.

```{r, cache=TRUE}
library(ggplot2)
library(tidyverse)

if (!file.exists('StormData.bz2')) {
	download.file('https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2', 'StormData.bz2')
}

raw <- read_csv('StormData.bz2')
```

### Health

First I decided to focus on events that caused the most damage to human health. I filtered the data by event type:

```{r, cache=TRUE}
fatbyevent <- raw %>% select(FATALITIES, EVTYPE) %>% group_by(EVTYPE) %>% 
	summarise(FATALITIES=sum(FATALITIES)) %>% arrange(desc(FATALITIES)) %>% 
	mutate(EVTYPE = ifelse(row_number() > 20, 'Other', EVTYPE)) %>% 
	group_by(EVTYPE) %>% summarise(FATALITIES=sum(FATALITIES))
```

Sorry for the enormous pipeline! Step by step, what I did is: selected the relevant data frame columns, grouped by event type, found total fatalities by group, rearranged and filtered to get the top 20 events by name (out of \~970 total events!), and re-summarised to get a nice and concise data frame.

Finally, I graphed the data and found the most harmful event type:

```{r, echo=TRUE, cache=TRUE}
g1 <- ggplot(fatbyevent, aes(EVTYPE, FATALITIES))
g1 + geom_col() + theme(axis.text.x = element_blank())
fatbyevent$EVTYPE[[which.max(fatbyevent$FATALITIES)]]
# Tornadoes cause the most damage to human health with a total of 
# 288661 recorded fatalities.
```

### Property

Now, I wanted to analyse which events caused the most damage to property. Again filtering by data type:

```{r, cache=TRUE}
dmgbyevent <- raw %>% select(PROPDMG, EVTYPE) %>% group_by(EVTYPE) %>% 
	summarise(PROPDMG=sum(PROPDMG)) %>% arrange(desc(PROPDMG)) %>% 
	mutate(EVTYPE = ifelse(row_number() > 20, 'Other', EVTYPE)) %>% 
	group_by(EVTYPE) %>% summarise(PROPDMG=sum(PROPDMG))
```

This does essentially the same thing as the previous data filtering (for total fatalities), but for property damage.

Graphing the data:

```{r, echo=TRUE, cache=TRUE}
g2 <- ggplot(dmgbyevent, aes(EVTYPE, PROPDMG))
g2 + geom_col() + theme(axis.text.x = element_blank())
dmgbyevent$EVTYPE[[which.max(dmgbyevent$PROPDMG)]]
# Again, tornadoes are the most harmful event type, this time by 
# property damage.
```

## Results

Tornadoes are terrible! They cause the most amount of fatalities and most amount of property damage out of any event type.
