---
title: "Leaflet Assignment"
author: "Alexander Ponomarev"
date: "2026-01-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## My R map in Leaflet

This assignment requires me to create a web page using R Markdown that features 
a map created with Leaflet. I must host this web page on RPubs. 

The point of this exercise is to briefly introduce myself to the Leaflet R
package.

First, let's load the data (available at https://data.a2gov.org/city-of-ann-arbor/air-quality-sensor-data):

```{r, message=FALSE}
library(tidyverse)
raw <- read_csv("ann_arbor_air_quality_hourly.csv")
```

## Making the plot

Now let's make a simple map for the data, I will only use entries from 2026
to keep the map small:

```{r, message=FALSE, warning=FALSE}
library(leaflet)
set.seed(2026-01-02)
filtered <- raw %>% filter(year(timestamp_hour_start)>2025)

df <- data.frame(lat=filtered$lat, lng=filtered$lon,
		 label=filtered$location_name, 
		 value=filtered$largest_pollutant_value)
		 
df %>% leaflet() %>% addTiles() %>% 
	addMarkers(label=~label,
		   popup=~value,
		   clusterOptions = markerClusterOptions())
```
