---
title: "Course Project"
author: "Me"
date: "2025-12-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The goal of this project is to predict the manner in which the subjects
performed the exercise. I will use the tidymodels package to build the model.

```{r}
library(tidyverse)
library(tidymodels)
raw_training <- read_csv('pml-training.csv')
raw_training %>% glimpse
```

## EDA

Now we can perform some EDA with this data. Note that the first column is 
simply the index (can cut it out of regression), last is the outcome.

```{r}
training <- raw_training %>% mutate(classe=as.factor(classe)) %>% 
	mutate(user_name=as.factor(user_name))
training$classe %>% levels
training %>% map(~ mean(is.na(.))) %>% glimpse
```

A lot of these variables have a very high NA rate and we can likely
get rid of them, they seem to be summary statistics. We can also remove
timestamps and other indexes.

## Building the model

We can move on to actually building the model now. We will make CV folds,
then make a recipe to normalize columns and remove nzv columns,
make a random forest model for each fold and finally run and collect the 
metrics.

```{r}
set.seed(123)

folds <- vfold_cv(training, v = 5, strata = classe)

rec <- recipe(classe ~ ., data=training) %>% 
	step_rm(`...1`, raw_timestamp_part_1, raw_timestamp_part_2, 
		cvtd_timestamp, new_window, num_window) %>% 
	step_filter_missing(all_predictors(), threshold=.95) %>% 
	step_impute_median(all_numeric_predictors()) %>% 
	step_normalize(all_numeric_predictors()) %>% 
	step_nzv(all_predictors())

mod <- rand_forest(trees = 200, mtry = 10, min_n = 5) %>%
	set_engine("ranger") %>%
	set_mode("classification")

wf <- workflow() %>% add_recipe(rec) %>% add_model(mod)

results <- fit_resamples(
	wf, resamples = folds,
	metrics = metric_set(accuracy, kap, mn_log_loss),
	control = control_resamples(save_pred = FALSE)
)

collect_metrics(results) %>% arrange(desc(mean))

final_fit <- fit(wf, data=training)
```

We can see that the model gave a high accuracy and good statistics for the 
kappa and mean log loss. So we will use this model.

```{r}
raw_testing <- read_csv('pml-testing.csv')
testing <- raw_testing %>% mutate(user_name=as.factor(user_name))
pred_class <- predict(final_fit, new_data = testing)
pred_class
```

These final predictions are what the project asks for.

Post-project notes: here are some things I think I can improve on. First off,
RF models don't require normalized columns. I was at first suspicious of the
high accuracy of the model, but the raw features and basically deterministic of
the outcome in this dataset (accelerations, velocities, orientations are
very strong predictors of if an exercise was done "correctly"). 